---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# metadid

[![tests](https://github.com/ben18785/metadid/actions/workflows/tests.yaml/badge.svg)](https://github.com/ben18785/metadid/actions)
[![Codecov test coverage](https://codecov.io/gh/ben18785/metadid/graph/badge.svg)](https://app.codecov.io/gh/ben18785/metadid)

**metadid** is an R package for Bayesian meta-analysis that synthesises treatment effects across studies with different designs: difference-in-differences (DiD), randomised controlled trials (RCT), and pre-post studies. It uses a hierarchical Stan model that accounts for design-specific information and heterogeneity across studies, with optional baseline normalisation to place outcomes on a common fractional scale.

## Installation

metadid depends on [cmdstanr](https://mc-stan.org/cmdstanr/) and [instantiate](https://CRAN.R-project.org/package=instantiate), which compile Stan models at package install time. Install them first if you haven't already:

```r
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()

install.packages("instantiate")
```

Then install metadid from GitHub:

```r
# install.packages("pak")
pak::pak("ben18785/metadid")
```

## Quick start

### 1. Simulate studies

`simulate_meta_did()` generates individual-level pre/post data for both arms across a set of studies from a known hierarchical model. `as_summary_did()` then aggregates this to the four-cell summary statistics (pre/post × control/treatment) that represent what is typically reported in a published study.

```{r simulate}
library(metadid)

sim <- simulate_meta_did(
  n_studies     = 20,
  true_effect   = -0.15,
  sigma_effect  = 0.03,
  baseline_mean = 0.45,
  rho           = 0.5,
  seed          = 42
)

studies <- as_summary_did(sim)
head(studies)
```

The true population treatment effect is `-0.15` on the raw scale, or approximately `r round(-0.15 / 0.45, 3)` after normalising by the baseline mean of `0.45`.

### 2. Fit the model via optimisation

`meta_did()` fits a hierarchical Bayesian model. Setting `method = "optimize"` finds the maximum a posteriori (MAP) estimate via L-BFGS — much faster than full MCMC and useful for quick iteration.

```{r fit, eval = FALSE}
fit <- meta_did(
  summary_data = studies,
  method       = "optimize",
  seed         = 42
)

print(fit)
```

```
#> Bayesian meta-analysis (metadid)
#> Studies: DiD = 20 | RCT = 0 | Pre-Post = 0 | DiD (change only) = 0 
#> Population treatment effect: -0.303  (MAP estimate, no uncertainty)
```

### 3. Inspect study-level estimates

`summary()` returns a data frame of population- and study-level parameters. With MAP optimisation, only point estimates are available (`sd`, `lo`, and `hi` are `NA`); use `method = "sample"` for full posterior uncertainty.

```{r summary, eval = FALSE}
summary(fit)
```

```
#>                       parameter       mean sd lo hi
#> 1         treatment_effect_mean -0.3032553 NA NA NA
#> 2           treatment_effect_sd  0.2621790 NA NA NA
#> 3  treatment_effect_did_summary -0.2626185 NA NA NA
#> 4  treatment_effect_did_summary -0.4008142 NA NA NA
#> 5  treatment_effect_did_summary -0.1788253 NA NA NA
#> ...
```

### 4. Fit the model via MCMC

For full posterior uncertainty, use the default `method = "sample"`:

```{r fit-mcmc, eval = FALSE}
fit_mcmc <- meta_did(
  summary_data  = studies,
  seed          = 42,
  iter_warmup   = 200,
  iter_sampling = 400
)

print(fit_mcmc)
```

```
#> Bayesian meta-analysis (metadid)
#> Studies: DiD = 20 | RCT = 0 | Pre-Post = 0 | DiD (change only) = 0 
#> Population treatment effect: -0.307  90% CI [-0.341, -0.273]
```

`summary()` now includes posterior standard deviations and credible intervals:

```{r summary-mcmc, eval = FALSE}
summary(fit_mcmc)
```

```
#>                       parameter        mean         sd          lo         hi
#> 1         treatment_effect_mean -0.30734301 0.02053925 -0.34109011 -0.2731060
#> 2           treatment_effect_sd  0.08387904 0.01614552  0.06170731  0.1124974
#> 3  treatment_effect_did_summary -0.23482038 0.02605055 -0.27537815 -0.1916570
#> 4  treatment_effect_did_summary -0.35087501 0.02539211 -0.39315193 -0.3085825
#> 5  treatment_effect_did_summary -0.22512112 0.02910410 -0.27397179 -0.1791810
#> ...
```
